Staged Abstract Interpreters

### 2019-09-09 paper/30_stageconc.tex
typo

To implement the staged concrete interpreter, we replay the steps from the
instantiation of the unstaged counterpart in Section \ref{unstaged_conc}.
However, now we use the @Rep@ type to annotate the value domains, environments and
stores, and redefine the staged version of monads and primitive operations.

\paragraph{Staged Monads}
We use the same monad stack structure as in the unstaged interpreter: a reader
monad with a state monad. The difference here is that, now the monads operate
on staged values.  For brevity, we call them \textit{staged monads}. In the
code snippet, we also use a @Rep@ prefix on the name of constructors and types
to differentiate them.
But it is important to note that the instances of @ReaderT@/@StateT@ are not
staged, and that the monadic computation, such as functions of @R => M[A]@, are
also not staged. Instead, the internal data that these monads operate on are
staged, i.e., values of @R@ and @A@ in the reader monad, and values of @S@ and
@A@ in the state monad.
The following code snippet shows the idea. We use \hl{light gray} to highlight
where @Rep@ types are added.
\begin{lstlisting}[escapechar=!]
  case class RepReaderT[M[_]: RepMonad, R, A](run: !\hl{Rep[R]}! => M[A]) {
    def flatMap[B](f: !\hl{Rep[A]}! => ReaderT[M, R, B]): RepReaderT[M, R, B] =
      RepReaderT(r => RepMonad[M].flatMap(run(r))(a => f(a).run(r))); ... }
  case class RepStateT[M[_]: RepMonad, S, A](run: !\hl{Rep[S]}! => M[(A, S)]) {
    def flatMap[B](f: !\hl{Rep[A]}! => StateT[M, S, B]): RepStateT[M, S, B] =
      RepStateT(s => RepMonad[M].flatMap(run(s))(as => f(as._1).run(as._2))); ... }
\end{lstlisting}

>>>>

To implement the staged concrete interpreter, we replay the steps from the
instantiation of the unstaged counterpart in Section \ref{unstaged_conc}.
However, now we use the @Rep@ type to annotate the value domains, environments and
stores, and define the staged version of monads and primitive operations.

\paragraph{Staged Monads}
We use the same monad stack structure as in the unstaged interpreter: a reader
monad with a state monad. The difference here is that, now the monads operate
on staged values.  For brevity, we call them \textit{staged monads}. In the
code snippet, we also use a @Rep@ prefix on the name of constructors and types
to differentiate them.
But it is important to note that the instances of @ReaderT@/@StateT@ are not
staged, and that the monadic computation, such as functions of @R => M[A]@, are
also not staged. Instead, the internal data that these monads operate on are
staged, i.e., values of type @R@ and @A@ in the reader monad, and values of type @S@ and
@A@ in the state monad.
The following code snippet shows the idea. We use \hl{light gray} to highlight
where @Rep@ types are added.
\begin{lstlisting}[escapechar=!]
  case class RepReaderT[M[_]: RepMonad, R, A](run: !\hl{Rep[R]}! => M[A]) {
    def flatMap[B](f: !\hl{Rep[A]}! => ReaderT[M, R, B]): RepReaderT[M, R, B] =
      RepReaderT(r => RepMonad[M].flatMap(run(r))(a => f(a).run(r))); ... }
  case class RepStateT[M[_]: RepMonad, S, A](run: !\hl{Rep[S]}! => M[(A, S)]) {
    def flatMap[B](f: !\hl{Rep[A]}! => StateT[M, S, B]): RepStateT[M, S, B] =
      RepStateT(s => RepMonad[M].flatMap(run(s))(as => f(as._1).run(as._2))); ... }
\end{lstlisting}

### 2019-09-09 paper/50_stageabs.tex
misc
typo

\paragraph{Primitive Operations} 
When deriving the staged concrete interpreter from its unstaged counterpart, we
notice that the recursive calls of the evaluation function @ev@ on the lambda
body are shifted from @ap_clo@ to @close@. In other words, we eagerly
specialize the interpreter and generate code every time when we reach a
$\lambda$ term, instead of lazily calling @ev@ when the applications happen.
Similarly, we use the same way to handle staged $\lambda$ terms.  Additionally,
in the staged abstract interpreter, we have to handle nondeterminism incurred
by the over-approximation of runtime behavior.  In the rest of this part, we
keep our roadmap and focus on discussing closure representations and function
applications for the staged version.

The @close@ method now is a mix of the \textit{staged} concrete version and
unstaged \textit{abstract} version. We first build a current-stage function @f@, which
takes four next-stage values, including the argument and latest store, and
@in/out@ caches additionally. Inside of @f@, we collapse the monadic value of
type @Ans@ to grounded values of type @Result@, by providing the desired
arguments, i.e., the new environment, new store, and two caches. The collapsing
of monadic values happens at the current-stage, so the invocation of @ev@ is unfolded
within staging. Finally, we generate a singleton set containing the compiled
next-stage closure @emit_compiled_clo(f)@, which is represented by an IR node
in LMS.
\begin{lstlisting}
  def emit_compiled_clo(f: (Rep[Value], Rep[Store], Rep[Cache], Rep[Cache])
                           => Rep[(Set[(Value,Store)], Cache)], λ: Lam, ρ: Exp[Env]): Rep[AbsValue]
  def close(ev: Expr => Ans)(λ: Lam, ρ: Rep[Env]): Rep[Value] = {
    val Lam(x, e) = λ
    val f: (Rep[Value],Rep[Store],Rep[Cache],Rep[Cache]) => Rep[(Set[(Value,Store)],Cache)] = {
      case (arg, σ, in, out) =>
        val α = alloc(σ, x)
        ev(e)(ρ + (unit(x) → α))(σ ⊔ Map(α → arg))(in)(out)
    }; Set[AbsValue](emit_compiled_clo(f, λ, ρ)) }
\end{lstlisting}

@@

In the staged interpreters, all instances of @Env@ and @Store@ are staged. The
data structures representing these components are treated as black-boxes, i.e.,
@Rep[Map]@, which means that the operations on a @Rep[Map]@ directly become
nex-stage code, and we do not inspect any further inside. As we identified when
introducing the generic interface, the keys of an @Env@ are string-represented
identifiers in the program, which are completely known statically. This
observation gives us a chance to further specialize the data structures for environment.
For example, let us assume that the @Map[K, V]@ is implemented as a hash-map.
If all the keys of type @K@ are all known statically, then the indices for
those keys can also be computed statically. Thus, the specialized map would be
an array of type @Array[Rep[V]]@, whose elements are next-stage values, and the
size of the array is known statically as the program has finite number of
identifiers. As result, an access to the map is staged into an access to the array
with the pre-computed index.

@@

\paragraph{Selective Caching} We observe that the two-fold co-inductive
caching is used for every recursive call in our abstract interpreter. But this
is not necessary and redundant when generating code for atomic
expressions such as literals or variables, because they always terminate.
Borrowing the idea from the partition of expressions used in A-Normal Form
$\lambda$-calculus \cite{Flanagan:1993:ECC:155090.155113}, we can use a
selective caching algorithm that does not generate caching code for atomic
expressions:
\begin{lstlisting}
  def fix_select: Expr => Ans = e => e match {
    case Lit(_) | Var(_) | Lam(_, _) => eval(fix_select)(e)
    case _ => fix_cache(e)
  }
\end{lstlisting}

>>>>

\paragraph{Primitive Operations} 
When deriving the staged concrete interpreter from its unstaged counterpart, we
notice that the recursive calls of the evaluation function @ev@ on the lambda
body are shifted from @ap_clo@ to @close@. In other words, we eagerly
specialize the interpreter and generate code every time when we reach a
$\lambda$ term, instead of lazily calling @ev@ when the applications happen.
Similarly, here we use the same way to handle staged $\lambda$ terms.  Additionally,
in the staged abstract interpreter, we have to handle nondeterminism incurred
by the over-approximation of runtime behavior.  In the rest of this part, we
keep our roadmap and focus on discussing closure representations and function
applications for the staged version.

The @close@ method now is a mix of the \textit{staged} concrete version and
unstaged \textit{abstract} version. We first build a current-stage function @f@, which
takes four next-stage values, including the argument and latest store, and
@in/out@ caches additionally. Inside of @f@, we collapse the monadic value of
type @Ans@ to grounded values of type @Result@, by providing the desired
arguments, i.e., the new environment, new store, and two caches. The collapsing
of monadic values happens at the current-stage, so the invocation of @ev@ is unfolded
within staging. Finally, we generate a singleton set containing the compiled
next-stage closure (@emit_compiled_clo@), which is represented by an IR node
in LMS.
\begin{lstlisting}
  def emit_compiled_clo(f: (Rep[Value], Rep[Store], Rep[Cache], Rep[Cache])
                           => Rep[(Set[(Value,Store)], Cache)], λ: Lam, ρ: Exp[Env]): Rep[AbsValue]
  def close(ev: Expr => Ans)(λ: Lam, ρ: Rep[Env]): Rep[Value] = {
    val Lam(x, e) = λ
    val f: (Rep[Value],Rep[Store],Rep[Cache],Rep[Cache]) => Rep[(Set[(Value,Store)],Cache)] = {
      case (arg, σ, in, out) =>
        val α = alloc(σ, x)
        ev(e)(ρ + (unit(x) → α))(σ ⊔ Map(α → arg))(in)(out)
    }; Set[AbsValue](emit_compiled_clo(f, λ, ρ)) }
\end{lstlisting}

@@

In the staged interpreters, all instances of @Env@ and @Store@ are staged. The
data structures representing these components are treated as black-boxes, i.e.,
@Rep[Map]@, which means that the operations on a @Rep[Map]@ directly become
nex-stage code, and we do not inspect any further inside. As we identified when
introducing the generic interface, the keys of an @Env@ are string-represented
identifiers in the program, which are completely known statically. This
observation gives us a chance to further specialize the data structures for environments.
For example, let us assume that the @Map[K, V]@ is implemented as a hash-map.
If all the keys of type @K@ are all known statically, then the indices for
those keys can also be computed statically. Thus, the specialized map would be
an array of type @Array[Rep[V]]@, whose elements are next-stage values, and the
size of the array is known statically as the program has finite number of
identifiers. As result, an access to the map is staged into an access to the array
with the pre-computed index.

@@

\paragraph{Selective Caching} We observe that the two-fold co-inductive
caching is used for every recursive call in our abstract interpreter. But this
is unnecessary and redundant when generating code for atomic
expressions such as literals or variables, because they always terminate.
Borrowing the idea from the partition of expressions used in A-Normal Form
$\lambda$-calculus \cite{Flanagan:1993:ECC:155090.155113}, we can use a
selective caching algorithm that does not generate caching code for atomic
expressions:
\begin{lstlisting}
  def fix_select: Expr => Ans = e => e match {
    case Lit(_) | Var(_) | Lam(_, _) => eval(fix_select)(e)
    case _ => fix_cache(e)
  }
\end{lstlisting}

### 2019-09-09 paper/60_cases.tex
typo

Specifically, \citeauthor{Boucher:1996:ACN:647473.727587} showed how to compile a
monovariant control-flow analysis \cite{Shivers:1991:SSC:115865.115884,
Shivers:1988:CFA:53990.54007} for continuation-passing style (CPS) programs.
The analyzer is a big-step control-environment abstract interpreter. 
By applying AC, we refactor every function in the analyzer such that it
return a closure just taking an environment argument.  Therefore the overhead
of traversing the AST of input program is eliminated.

@@

On the other side, our approach works equivalently through staging: the ASTs
are static, whereas stores are dynamic. Therefore, we annotate the type @Store@
with @Rep@ (Figure \ref{compare_ac_sai} (right)). In fact, the only changes of
our approach using the LMS framework are the types of @Store@ to @Rep[Store]@,
since they will be known at the next stage. There is no need for other changes.
The generated programs consist entirely of looks-up and updating operations
w.r.t the store.  The functions at the current stage are eliminated in the
residual program; therefore, we do not generate higher-order functions, which
provides additional performance improvement. Nevertheless, the LMS framework
provides support for necessary data structures, such as @Map@.  We consider
these efforts reusable and completely orthogonal to the implementation of the
analysis.

>>>>

Specifically, \citeauthor{Boucher:1996:ACN:647473.727587} showed how to compile a
monovariant control-flow analysis \cite{Shivers:1991:SSC:115865.115884,
Shivers:1988:CFA:53990.54007} for continuation-passing style (CPS) programs.
The analyzer is a big-step control-environment abstract interpreter. 
By applying AC, we refactor every function in the analyzer such that it
returns a closure taking an environment argument.  Therefore the overhead
of traversing the AST of input program is eliminated.

@@

On the other side, our approach works equivalently through staging: the ASTs
are static, whereas stores are dynamic. Therefore, we annotate the type @Store@
with @Rep@ (Figure \ref{compare_ac_sai} (right)). In fact, the only changes of
our approach using the LMS framework are the types of @Store@ to @Rep[Store]@,
since they will be known at the next stage. There is no need for other changes.
The generated programs consist entirely of looks-up and updating operations
w.r.t the store.  The functions at the current stage are eliminated in the
residual program; therefore, we do not generate higher-order functions, which
provides additional performance improvement. Nevertheless, the LMS framework
provides support for necessary data structures, such as for @Map@.  We consider
these efforts reusable and completely orthogonal to the implementation of the
analysis.
